---
title: "Initial Plotting"
author: "Brandon Sherman"
date: "July 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We've already downloaded the relevant data and munged it into tibbles
thanks to the wonderful `spotifyr` package.  See `0-DownloadRush.Rmd` for all the 
dirty work.

In this initial notebook, we want to understand what musical features
differentiate the various recording periods of Rush.

```{r prep}
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(ggrepel)
library(ggforce)
library(patchwork)
library(here)
library(stringr)
library(tidyr)
library(glue)
library(extrafont)
library(forcats)

source(here("lib", "vars.R"))

rush_albums <- readRDS(RUSH_ALBUMS)
```

First let's plot each variable over time.  To do this, we need to
first gather all of the musical feature columns.  Then we'll make a
facetted plot.

There's one issue though, which is that we have 165 tracks.  That's a
lot of dots for a time series!  We'll calculate a weighted average of
each of these variables at the album level.  The weighting will be by
track duration because Rush has many long songs that dominate most of
an album.  The title track of 2112, which is 20 minutes long, should not 
be weighted the same as A Passage to Bangkok, which is 3 minutes long.

```{r plot_data_prep}
album_duration_dat <- rush_albums %>%
    group_by(album_name) %>%
    summarize(album_duration_ms = sum(duration_ms))

album_stats <- rush_albums %>%
    group_by(album_name) %>%
    gather(feature, value,
           danceability, energy, acousticness, instrumentalness, valence, tempo) %>%
    inner_join(album_duration_dat, by = "album_name") %>% 
    mutate(track_weight = duration_ms / album_duration_ms,
           weighted_value = track_weight * value) %>%
    group_by(album_name, album_release_date, feature) %>% 
    summarize(value = mean(value),
              weighted_value = sum(weighted_value)) %>%
    ungroup() %>% 
    mutate(feature = Hmisc::capitalize(feature)) %>% ## acousticness -> Acousticness
    arrange(album_release_date)
```

```{r make_plot}
## Thanks to
## http://www.2112.net/rushfonts/
## for the font!
for (i in seq_len(3)) {
    album_stats_plots <- album_stats %>%
        rename(Feature = feature) %>% 
        ggplot(aes(x = album_release_date, y = weighted_value, label = album_name)) +
        geom_line(aes(color = Feature)) +
        geom_label_repel(alpha = 0.6, size = 2) +
        facet_wrap_paginate(~ Feature, scales = "free", ncol = 1, nrow = 2, page = i) +
        labs(title = "Musical development of Rush over time"
             ) +
        theme(plot.title = element_text(family = "Working Man", size = 22),
              legend.position = "none") +
        xlab("Release Date") +
        ylab("Value")

    album_stats_plots %>%
        ggsave(here("figures", glue("feature_plot_{i}.png")), ., width = 7, height = 7, units = "in")
}

```


Here's what we see:

* Moving Pictures is not only the most instrumental Rush album, but it
  marks a peak at which Rush's music got less instrumental.  As they
  added more synthesizers to fit in with their 80s contemporaries, Rush 
  also added more vocals to the mix.  However Rush has never been close to 
  an instrumental band, as we see no "Instrumentalness" values above 0.2  
  For reference, the highest possible "Instrumentalness" value is 1.
  
* Caress of Steel is Rush's most acoustic album, and also has among
  the lowest danceability and energy scores.
  
* Power Windows, which is dominated by synthesizers and fast tracks,
  has high energy and low acousticness.
  
* Rush's music has gotten slower over time.  It's interesting how
after Power Windows, when Rush decided that they didn't want to be a
progressive rock band dominated by synthesizers, Rush's music quickly 
started dropping in average tempo over the course of their next four 
albums, reaching 120BPM on Counterparts and staying approximately 
there until Clockwork Angels.

There is one data point that I am very curious about.  Why does Fly By
Night have such a high instrumentalness score when all of its tracks
have lyrics?  Let's take a look at the individual track values for the
album.

```{r fly_by_night}
rush_albums %>%
    filter(album_name == "Fly By Night") %>%
    select(track_name, instrumentalness, duration_ms)
```

My suspicion is that it's being driven by By-Tor and the Snow Dog,
which is the longest track on the album and has long instrumental
sections.  Similarly, Anthem and Best I Can have lyrics but lean
instrumental.  Anthem has no vocals until 54 seconds into a four
minute song.  Best I Can has similar instrumental sections.



What's the distribution of song lengths per album?


```{r}
rush_albums_for_circle_plot <- rush_albums %>%
    group_by(album_name) %>%  ## Order and label the tracks for each album descendingly by length
    arrange(desc(duration_ms)) %>%
    mutate(rn = row_number()) %>%
    ungroup() %>% 
    mutate(minutes = floor(duration_ms / 6e4)) %>%
    mutate(seconds = floor(duration_ms / 1000) %% 60) %>%
    mutate(label = glue("{track_name}\n{minutes}:{seconds}")) %>% 
    arrange(album_release_date)

rush_albums_for_circle_plot %>% 
    ggplot(aes(x = fct_reorder(album_name, album_release_date, .desc = TRUE), ## TRUE because coord_flip
               y = rn, size = duration_ms, color = album_name)) +
    geom_point() +
    coord_flip() +
    ggtitle("Track lengths by album") +
    xlab("") +
    theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
    guides(color = FALSE) +
    annotate("text", x = "2112", y = 1, label = "2112\n20:34", size = 2) +
    annotate("text", x = "Hemispheres", y = 1, label = "Cygnus X-1\n18:08", size = 2) +
    annotate("text", x = "Caress Of Steel", y = 1.1, label = "The Fountain of\nLamneth\n19:57", size = 2)

## bars

## Thanks to
## https://stackoverflow.com/a/50173616
## for how to make mm:dd labels

for (i in seq_len(5)) {
    
    duration_plot <- rush_albums_for_circle_plot %>%
        mutate(duration_ms = as.difftime(duration_ms / 1000, units = "secs")) %>%
        ## Truncate some long track names
        mutate(track_name = if_else(grepl("2112", .$track_name), "2112", track_name),
               track_name = if_else(grepl("La Villa Strangiato", .$track_name), "La Villa Strangiato", track_name)) %>%
        mutate(album_name = fct_reorder(album_name, album_release_date)) %>% 
        ggplot(aes(x = fct_reorder(track_name, duration_ms),
                   y = duration_ms, color = album_name, fill = album_name)) +## TRUE because coord_flip
        geom_col(width = 0.8) +
        scale_y_time(labels = function(l) strftime(l, "%M:%S")) +
        facet_wrap_paginate(~ album_name, scales = "free_y", drop = TRUE, ncol = 1, nrow = 4, page = i) +
        ggtitle("Rush Track Lengths by Album") +
        ylab("Track Length") +
        xlab("") +
        coord_flip() +
        theme(plot.title = element_text(family = "Working Man", size = 22),
              legend.position = "none")

    duration_plot %>% 
        ggsave(here("figures", glue("duration_plot_{i}.png")), ., height = 7, width = 7)
}

```

## Output

```{r output}
saveRDS(album_stats, ALBUM_STATS_OUTPUT)
```
