---
title: "Clustering"
author: "Brandon Sherman"
date: "July 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Now that we've looked at Rush's musical development over time, let's
do some hierarchical clustering based on music features.  This is
heavily inspired by, if not plagiaried, directly from [Alyssa
Goldberg's similar analysis on David Bowie albums](https://twitter.com/WireMonkey/status/1009915034246565891?s=19).

```{r prep}
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(here)
library(rpart)
library(dendextend) ## for colors
library(circlize) ## one of the only times where circles are nice

source(here("lib", "vars.R"))

album_stats <- readRDS(ALBUM_STATS)
```

First of all, our data is in long format where each row is "musical
feature of an album".  We need it to be in wide format, where each row
is "all musical features for an album".  To do this, let's use
`tidyr::spread`.  After all, it's the opposite operation of
`tidyr::gather`, which we utilized when prepping this data.

We'll make two frames: One where the values have been weighted by
track length, and another where they haven't been.  Since `hclust` needs 
a matrix of values, we'll also do that transformation here.

Since the process to make both data frames is virtually the same, I'll
write a function that uses `dplyr`'s new quasiquotation features.  I'd
like to pass multiple columns for `col_to_drop`, but sadly this seems like 
more trouble than it's worth right now.

```{r data_prep}
prep_df_for_cluster <- function(dat, val_col, col_to_drop) {
    ## Given a data frame, a value column, and a bunch of columns to drop,
    ## return a data.frame appropriate for clustering.

    val_col <- enquo(val_col)
    col_to_drop <- enquo(col_to_drop)
    
    dat %>%
        select(-!!col_to_drop, -album_release_date) %>% 
        spread(feature, !!val_col) %>%
        ## mutate_if(is.numeric, scale) %>%
        mutate(Tempo = scale(Tempo)) %>% 
        column_to_rownames("album_name") %>%
        as.data.frame()
}    

rush_albums_for_weighted_cluster <- prep_df_for_cluster(album_stats, weighted_value, value)
rush_albums_for_unweighted_cluster <- prep_df_for_cluster(album_stats, value, weighted_value)
```


## Clustering

```{r clustering}
dend <- as.dendrogram(hclust(dist(rush_albums_for_weighted_cluster), method = "single")) %>%
    color_branches(k = 5) %>%
    color_labels(k = 5)

circlize_dendrogram(dend, labels_track_height = 0.4, dend_track_height = 0.4)
dev.copy(png, here("figures", "weighted_cluster_plot.png"))
dev.off()


unweighted_dend <- as.dendrogram(hclust(dist(rush_albums_for_unweighted_cluster), method = "complete")) %>%
    color_branches(k = 4) %>%
    color_labels(k = 4)

circlize_dendrogram(unweighted_dend, labels_track_height = 0.4, dend_track_height = 0.4)
dev.copy(png, here("figures", "unweighted_cluster_plot.png"))
dev.off()
```

Interestingly enough, using the *unweighted* dendrogram leads to
better clusters than the weighted one.  I suspect it's because we're
doing additional transformations on transformed data.

In the unweighted dendrogram, we see the following clusters:

* Purple:  hard rock with some weird effects
* Red:  80s albums with lots of synthesizers
* Yellow-Green:  Early hard rock
* Green:  Later hard rock

Followup to last week's work tracking Rush's musical development over
time. Here's a hierarchical clustering of Rush albums. Albums in the
same color are similar to each other.


## Robustness Checking

Now suppose that Rush decided to reunite release a brand new album
that's considerably more poppy and Spanish than what they put out in
the past four decades.  For the same of argument, let's call this new
track "Despacito".  I'm going to query Despacito, add it to Rush's
discography, and see how it influences the clustering.  If these are 
good clusters, then they should be virtually the same with or without 
Despacito.

I'll also do it for Africa by Toto because why not?


```{r}
library(spotifyr)

get_single_track_features <- function(track_name, slice_num) {
    
    get_tracks(track_name) %>%
        slice(slice_num) %>% 
        get_track_audio_features() %>%
        select(danceability, energy, acousticness, instrumentalness, valence, tempo) %>%
        mutate(album_name = track_name) %>%
        gather(feature, value, acousticness, danceability, energy, instrumentalness, tempo, valence) %>%
        mutate(feature = Hmisc::capitalize(feature)) %>%
        mutate(weighted_value = value)
}    

despacito <- get_single_track_features("Despacito", 2)
africa <- get_single_track_features("Africa", 1) %>% mutate(album_name = "Africa by Toto")


album_stats_with_new_tracks_for_unweighted_clustering <- album_stats %>%  ## Phew!
    bind_rows(africa) %>%
    bind_rows(despacito) %>% 
    prep_df_for_cluster(value, weighted_value)


unweighted_dend_with_new_tracks <- as.dendrogram(hclust(dist(album_stats_with_new_tracks_for_unweighted_clustering), method = "complete")) %>%
    color_branches(k = 6) %>%
    color_labels(k = 6)

circlize_dendrogram(unweighted_dend_with_new_tracks, labels_track_height = 0.4, dend_track_height = 0.4)
```

Despacito and Africa are their own cluster.  Which makes sense because
I've never been to a club that played Tom Sawyer by Rush.
